---
title: "PLeon_Unit6CaseStudy"
author: "Paola Leon"
date: "June 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The purpose of this document is to analyze the Gross Domestic Product (GDP) among economies with confirmed GDP estimates only. The data includes observations for the year of 2012 (mainly) but can include data for 2011 or 2010 in some cases. The study focuses in understanding the different GDP rankings, quantiles and the correlation that exists (if any) with the Income Groups defined in the World Bank EdStats data.


Original data sources can be found in the links below:

http://data.worldbank.org/data-catalog/GDP-ranking-table

http://data.worldbank.org/data-catalog/ed-stats


#Recording the session info

```{r}
sessionInfo()
```

#Set the working directory

```{r}
setwd("/Users/paolaleon/Desktop/MSDS 6303 Doing Data Science/Unit6CaseStudy")
getwd()
```

#Download datasources.

1. If needed first install and load the "downloader" packages using the following code:

install.packages("downloader")

library("downloader")

2. Download the raw data from the internet by typing:

download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv", destfile="GDP.csv")

download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv", destfile="EDU.csv”)

3. Verify that the files exist in the working directory (downloaded succesfully):

```{r}
list.files()
```

#Analysis of raw data: GDP

```{r}
rawGDP <- read.csv("GDP.csv", stringsAsFactors=FALSE, header=FALSE)
```

General description of the dataframe: number of observations, number of variables, datatypes, and existance of NAs.

```{r}
str(rawGDP)
```

A snapshot of the dataframe: top six rows

```{r}
head(rawGDP)
```

A snapshot of the dataframe: last six rows

```{r}
tail(rawGDP)
```

Identify the names of the different variables (column headers):

```{r}
names(rawGDP)
```

Identify the dimensions of the dataframe (number of observations and variables):

```{r}
dim(rawGDP)
```

Calculate dataframe summaries for both continuos(numeric) and categorical fields:

```{r}
summary(rawGDP)
```

#Analysis of raw data: EDU

```{r}
rawEDU <- read.csv("EDU.csv", stringsAsFactors=FALSE, header=FALSE)
```

General description of the dataframe: number of observations, number of variables, datatypes, and existance of NAs.

```{r}
str(rawEDU)
```

A snapshot of the dataframe: top six rows

```{r}
head(rawEDU)
```

A snapshot of the dataframe: last six rows

```{r}
tail(rawEDU)
```

Identify the names of the different variables (column headers):

```{r}
names(rawEDU)
```

Identify the dimensions of the dataframe (number of observations and variables):

```{r}
dim(rawEDU)
```

Calculate dataframe summaries for both continuos(numeric) and categorical fields:

```{r}
summary(rawEDU)
```

#Tidyng Data EDU

1. If needed install and load the following packages:

install.packages("tidyr")

library("tidyr")

install.packages("dplyr")

library("dplyr")

2. Create a new dataframe for tyding data so that raw data stays untouched.

```{r}
dataEDU <- rawEDU
```

3. Start by verifying again the inital structure of the data to get a sense of what needs to be done and cleaned.

```{r}
str(dataEDU)
```

4. Rename the variables to meaninful labels

```{r}
names(dataEDU) <- c("CountryCode", "CountryName", "IncomeGroup", "Region", "Lending category", "Other groups", "Currency", "Population", "Notes", "National accounts base year", "National accounts reference year", "System of National Accounts", "SNA price valuation", "Alternative conversion factor", "PPP survey year", "Balance of Payments Manual in use", "External debt Reporting status", "System of trade", "Government Accounting concept", "IMF data dissemination standard", "Source of most recent Income and expenditure data", "Vital registration complete", "Latest agricultural census", "Latest industrial data", "Latest trade data", "Latest water withdrawal data", "2-alpha code", "WB-2 code", "Table Name", "Short Name")
```

5. Only select those variables that will be relevant in our future data analysis: CountryCode, CountryName and IncomeGroup.

```{r}
datasetEDU <- dataEDU[c(1:3)]
```

6. For variable CountryCode replace the value "CountryCode" for nothing to eliminate invalid value.

```{r}
datasetEDU$CountryCode <- gsub("CountryCode", "", datasetEDU$CountryCode)
```

7. Also for variable CountryName replace the value "Long Name" for nothing to eliminate invalid value.
```{r}
datasetEDU$CountryName <- gsub("Long Name", "", datasetEDU$CountryName)
```

8. Check how the data is looking so far.

```{r}
str(datasetEDU)
```

9. One more replace of an invalid value.

```{r}
datasetEDU$IncomeGroup <- gsub("Income Group", "", datasetEDU$IncomeGroup)
```

10. See a snapshot of the tidy data:
```{r}
head(datasetEDU)
```

11. Select only relevant rows (skip empty rows):
```{r}
datasetEDU <- datasetEDU[2:235,]
```

12. Check one more time to make sure we have the data ready and clean

```{r}
str(datasetEDU)
```

#Tidyng Data GDP

1. Create a new dataframe for tyding data so that raw data stays untouched.

```{r}
dataGDP <- rawGDP
```

2. Start by verifying again the inital structure of the data to get a sense of what needs to be done and cleaned.

```{r}
str(dataGDP)
```

3. Select only those columns relevant for our future data analysis:
```{r}
datasetGDP <- dataGDP[c(1,2,5)]
```

4. Rename variables into more suitable names:

```{r}
names(datasetGDP) <- c("CountryCode", "Ranking", "MillionsUSD")
```

5. See how the data is looking so far:

```{r}
str(datasetGDP)
head(datasetGDP)
```

6. Remove invalid values
```{r}
datasetGDP$MillionsUSD <- gsub("\\(millions of", "", datasetGDP$MillionsUSD)
datasetGDP$MillionsUSD <- gsub("US dollars\\)", "", datasetGDP$MillionsUSD)
datasetGDP$Ranking <- gsub("Gross domestic product 2012", "", datasetGDP$Ranking)
datasetGDP$Ranking <- gsub("Ranking", "", datasetGDP$Ranking)
```

7. Verify again:
```{r}
head(datasetGDP)
```

8. Skip empty rows:
```{r}
datasetGDP <- datasetGDP[6:331,]
```

9. For variable MillionsUSD prepare for datatype conversion:
```{r}
datasetGDP$MillionsUSD <- gsub(",", "", datasetGDP$MillionsUSD)
```

10. For variable MillionsUSD convert to datatype integer:
```{r}
datasetGDP$MillionsUSD <- as.integer(datasetGDP$MillionsUSD)
```

11. Calculate summaries and identify number or NAs (if any):
```{r}
summary(datasetGDP)
```

Number of NA's = 122

12. Identify anomalies with CountryCode
```{r}
table(datasetGDP$CountryCode)
```

13. Discard invalid rows (rows with empty CountryCode)
```{r}
datasetGDP <- datasetGDP[datasetGDP$CountryCode != "", c(1,2,3)]
```

14. For variable Ranking convert to datatype integer:
```{r}
datasetGDP$Ranking <- as.integer(datasetGDP$Ranking)
```

15. Check one more time to make sure we have the data ready and clean

```{r}
str(datasetGDP)
```

#Merge data
**Match the data based on the country shortcode. How many of the IDs match?**

1. See how many observations we have in both datasets:  datasetGDP and datasetEDU
```{r}
dim(datasetGDP)
dim(datasetEDU)
```

2. Merge the datasets by country shortcode (variable CountryCode):
```{r}
datamerge <- merge(datasetEDU, datasetGDP, "CountryCode")
```

3. See how many observations matched: 224
```{r}
dim(datamerge)
```

**Sort the data frame in ascending order by GDP (so United States is last).**

1. First eliminate rows with no GDP

```{r}
datamerge <- subset(datamerge, !is.na(MillionsUSD))
```

2. Sort in ascending order by GDP:

```{r}
library("dplyr")
datamerge <- arrange(datamerge, MillionsUSD)
```

**What is the 13th country in the resulting data frame?**

```{r}
head(datamerge,13)
```

The 13th country in the list is: KNA - St. Kitts and Nevis

**What are the average GDP rankings for the "High income: OECD" and "High income:
nonOECD" groups?**

1. Choose only the observations with IncomeGroup = "High income: OECD" and calculate its mean

```{r}
HighIncomeOECD <- datamerge[datamerge$IncomeGroup == "High income: OECD", c(2,5)]
HighIncomeOECD
summary(HighIncomeOECD$MillionsUSD)
```

2. Choose only the observations with IncomeGroup = "High income: nonOECD" and calculate its mean

```{r}
HighIncomeNonOECD <- datamerge[datamerge$IncomeGroup == "High income: nonOECD", c(2,5)]
HighIncomeNonOECD
summary(HighIncomeNonOECD$MillionsUSD)
```


The average GDP for the Income Group "High income: OECD" is: 1,484,000

The average GDP for the Income Group "High income: nonOECD" is: 1,04,300

#Plot the GDP for all of the countries. 

**Use ggplot2 to color your plot by Income Group.**

1. If needed install and load the "ggplot2" package as follows:

install.packages("ggplot2")

library("ggplot2")

```{r}
library("ggplot2")
```

2. Get data ready for plot (discard unwanted rows such as summaries and rows with empty values):

```{r}
dataplot <- datamerge[1:198,]
dataplot <- dataplot[dataplot$IncomeGroup != "", c(1,2,3,4,5)]
```

3. Plot using qplot syntax
```{r}
qplot(MillionsUSD, CountryName, data=dataplot, color=IncomeGroup, geom=c("point"))
```

4. Plot using ggplot syntax
```{r}
ggplot(data=dataplot, aes(x=MillionsUSD,y=CountryName, color=IncomeGroup))+geom_point()
```

Interpretation of the Plot: United States of America was by far the country with the highest Gross Domestic Product for 2012. Also the countries classified in the Lower Middle Income Group and the Upper Middle Income Group seem to be the majority within the countries for which the GDP for 2012 was analyzed.


#Cut the GDP ranking into 5 separate quantile groups. 

1. Create a new dataframe for analysis and review its structure.

```{r}
dataanalysis <- dataplot
str(dataanalysis)
```

2. Get the mean and standard deviation of the Ranking variable for further calculations of quantiles
```{r}
summary(dataanalysis$Ranking) 
sd(dataanalysis$Ranking)
```

Mean:95.31

SD:55.06931

3. Calculate quantiles, e.g. the value below which 20 percent of the contries is Ranked
```{r}
Below20Quantile <- qnorm(0.20, 95.31, 55.06931)
Below20Quantile
Below40Quantile <- qnorm(0.40, 95.31, 55.06931)
Below40Quantile
Below60Quantile <- qnorm(0.60, 95.31, 55.06931)
Below60Quantile
Below80Quantile <- qnorm(0.80, 95.31, 55.06931)
Below80Quantile
Below99Quantile <- qnorm(.99, 95.31, 55.06931)
Below99Quantile
dataBelow20Quatile <- dataanalysis[dataanalysis$Ranking <= Below20Quantile, c(1,2,3,4,5)]
dataBelow40Quatile <- dataanalysis[dataanalysis$Ranking <= Below40Quantile, c(1,2,3,4,5)]
dataBelow60Quatile <- dataanalysis[dataanalysis$Ranking <= Below60Quantile, c(1,2,3,4,5)]
dataBelow80Quatile <- dataanalysis[dataanalysis$Ranking <= Below80Quantile, c(1,2,3,4,5)]
dataBelow99Quatile <- dataanalysis[dataanalysis$Ranking <= Below99Quantile, c(1,2,3,4,5)]
```

#Make a table versus Income.Group

```{r}
table(dataanalysis$Ranking, dataanalysis$IncomeGroup)
```

**How many countries are Lower middle income but among the 38 nations with highest GDP?**

Per the counts in the table above: 5


#Conclusion

1. I really enjoyed working in this case study. It provided a great opportunity to practice many of the R commands we have learned in previous weeks. The instructions also served as a guide to the sequence of events that ultimately allowed for the natural development of the case having a dataflow in the order of raw data -> tidy data -> analyzed data.

2. The 'technical' challenges involved in the resolution of the ggplot2 and quantiles questions required additional time and effort but it was gratifying researching and trying more regarding these topics.

3. Here is also a short list that I would like to keep in mind for a future reference:
- Always keep raw data untouched, saved in a separate variable.
- Always maintain sequential variables to hold the progress done at a certain point in time.
- Always before attempting to convert into numeric data type, make sure the values to be converted have discarded any characters such as , $, etc. Otherwise NAs could result upon data type conversion.
- When in doubt and as an optional practice, it is a good idea to open the file in other applications such as MS Excel or Text Edit to verify column headers and the structure of the file.
- Also as part of tidying data, it is recommended to constantly verify the work done over the data to make sure the resultset is as expected. Commands such as head, tail, str, dim, names should be used often.
- It also resulted valuable discarding empty rows while preparing the data for plotting.









#Trying to have some fun

1. If needed install and load the following package:

install.packages("googleVis")

library("googleVis")

```{r}
library("googleVis")
```

2. Use the Google gvisGeoChart to map the data based on IncomeGroup

```{r echo-FALSE, message=FALSE, results='asis'}
GDPMap <- gvisGeoChart(dataplot, dataplot$CountryCode, dataplot$IncomeGroup) 
plot(GDPMap)
```




___

*Gandrud Christopher, Reproducible research with R and RStudio, CRC Press, 2015 Edition*

*Teutonico Donato, ggplot2 Essentials, Packt Publishing, June 25, 2015*

*Stowell Sarah, Using R for Statistics, Apress, June 29, 2014*



